Exploring Red Wine Quality by Ludmila Levkova
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(GGally)
library(dplyr)
library(bitops)
library(RCurl)
library(memisc)
library(lattice)
library(MASS)
library(RColorBrewer)
library(e1071)
library(randomForest)
library(neuralnet)

```

```{r echo=FALSE, Load_the_Data}
factor2numeric <- function(f)
{
   if(!is.factor(f)) stop("the input must be a factor")
   as.numeric(levels(f))[as.integer(f)]
}
# Load the Data
rw<-read.csv2("winequality/winequality-red.csv")
for (i in 1:11){
rw[,i]<-factor2numeric(rw[,i])
}

#Normalize
rw_n<-rw
for (i in 1:11){
    rw_n[,i]<- (rw_n[,i]-mean(rw_n[,i]))/sd(rw_n[,i])
}

#training set
n<-1599
ntrain<-round(n*0.8)
ntest<-n-ntrain

```
```{r echo=FALSE, Models}
#linear regression
set.seed(1501112015)
N<-10
ACC<-c(rep(0,N))
RMSE<-c(rep(0,N))

for (i in 1:N){
  tindex <- sample(n,ntrain) # indices of training samples
  rw_train <- rw_n[tindex,]
  rw_test <- rw_n[-tindex,]
m1 <- lm(I(quality) ~ alcohol, data = rw_train)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates )
m4 <- update(m3, ~ . + total.sulfur.dioxide )
m5 <- update(m4, ~ . + chlorides)
m6 <- update(m5, ~ . + fixed.acidity)
m7 <- update(m6, ~ . + citric.acid)
m8 <- update(m7, ~ . + density)
m9 <- update(m8, ~ . + pH)
m10<- update(m9,~ . + residual.sugar)
m11<- update(m10,~. + free.sulfur.dioxide)
mtable(m1,m2, m3, m4, m5, m6, m7, m8,m9,m10,m11)
res<-predict(m4,rw_test[,-12])
RMSE[i]<-sqrt(crossprod(res-rw_test$quality)/ntest)
tab<-table(pred=round(res),true = rw_test[,12])
ACC[i]<-classAgreement(tab, match.names=T)$diag
}
mean(RMSE); sd(RMSE)
mean(ACC); sd(ACC)

#LR Ridge
Ridge<-lm.ridge(I(quality) ~., data = rw_train, lambda=seq(0,100,0.001))
select(Ridge)
R1<-lm.ridge(I(quality) ~., data = rw_train, lambda=44.1)
res_r<-coef(R1)[1]
for (i in 2:11){
   res_r<-res_r+coef(R1)[i]*rw_test[,i]
}
MSPE_r<-sum((res_r - rw_test$quality)^2)/ntest
MSPE_r
      

#SVM
rw_n1<-rw_n
rw_n1$quality <- as.factor(rw_n1$quality)
tindex <- sample(n,ntrain) # indices of training samples
rw_train <- rw_n1[tindex,]
rw_test <- rw_n1[-tindex,]
wts = ntrain/table(rw_train$quality)
tune.svm(quality~.,data=rw_train, gamma = 2^(-2:1), cost = 2^(1:6), class.weights = wts) #gamma=1 cost=4
tune.svm(quality~alcohol+sulphates+total.sulfur.dioxide+volatile.acidity,data=rw_train, gamma = 2^(-1:1), cost = 2^(2:6))#gamma=1 cost=4

ACC1<-c(rep(0,10))
for (i in 1:10){
  tindex <- sample(n,ntrain) # indices of training samples
  rw_train <- rw_n[tindex,]
  rw_test <- rw_n[-tindex,]  
#mysvm1<-svm(quality~alcohol, data=rw_train[,], cost=4)
#mysvm2<-update(mysvm1, ~.+ volatile.acidity)
#mysvm3<-update(mysvm2, ~.+ sulphates)
#mysvm4<-update(mysvm3, ~.+ total.sulfur.dioxide)
#mysvm5<-update(mysvm4, ~.+ chlorides)
#mysvm6<-update(mysvm5, ~.+ pH)
#mysvm7<-update(mysvm6, ~.+ free.sulfur.dioxide)
#mysvm8<-update(mysvm7, ~.+ density)
#mysvm9<-update(mysvm8, ~.+ fixed.acidity)
#mysvm10<-update(mysvm9, ~.+ citric.acid)
#mysvm11<-update(mysvm10, ~.+ residual.sugar)
#plot(cmdscale(dist(rw[,-12])), col =as.integer(rw[,12]))
mysvm1a<-svm(quality~alcohol+sulphates+total.sulfur.dioxide+volatile.acidity, data=rw_train[,], cost=4, gamma=1)
#mysvm1a<-svm(quality~.,data=rw_train[,], cost=4, gamma=1)
res1 <- predict(mysvm1a, newdata=rw_test)
MSPE1[i]<-crossprod(res1 - as.numeric(rw_test$quality))/ntest
tab1<-table(pred = res1, true = rw_test[,12])
ACC1[i]<-classAgreement(tab1,match.names=T)$diag
}
mean(ACC1);sd(ACC1)
mean(MSPE1); sd(MSPE1)



# regression tree/random forest
N<-10
ACC2<-c(rep(0,N))
RMSE2<-c(rep(0,N))
set.seed(15011120)
for (i in 1:N){
  tindex <- sample(n,ntrain) # indices of training samples
  rw_train <- rw_n[tindex,]
  rw_test <- rw_n[-tindex,] 
#reg<-randomForest(quality~., data=rw_train, method="class")
reg<-randomForest(quality~alcohol+sulphates+total.sulfur.dioxide+volatile.acidity, data=rw_train) 
#reg<-rpart(quality~., data=rw_train, method="class")
#reg<-rpart(quality~alcohol+sulphates+total.sulfur.dioxide+volatile.acidity, data=rw_train, method="class")
res2 <- predict(reg, newdata=rw_test)
RMSE2[i]<-sqrt(crossprod(res2 - as.numeric(rw_test$quality))/ntest)
tab2<-table(pred = round(res2), true = rw_test[,12])
ACC2[i]<-classAgreement(tab2,match.names=T)$diag
}

mean(ACC2);sd(ACC2)
mean(RMSE2); sd(RMSE2)
#p<-predict(reg, newdata1)

# neural network
ACC3<-c(rep(0,100))
MSPE3<-c(rep(0,100))
for (i in 1:100){
  tindex <- sample(n,ntrain) # indices of training samples
  rw_train <- rw_n[tindex,]
  rw_test <- rw_n[-tindex,] 
#net<-neuralnet(quality~volatile.acidity, data=rw_train, hidden=10, threshold=0.1)
net<-neuralnet(quality~alcohol+sulphates+total.sulfur.dioxide+volatile.acidity, data=rw_train, hidden=5, threshold=0.01) 
#reg<-rpart(quality~., data=rw_train, method="class")
#reg<-rpart(quality~alcohol+sulphates+total.sulfur.dioxide+volatile.acidity, data=rw_train, method="class")
res3 <- compute(net, rw_test[,c("alcohol","sulphates","total.sulfur.dioxide","volatile.acidity")])
MSPE3[i]<-crossprod(round(res3$net.result) - as.numeric(rw_test$quality))/ntest
tab3<-table(pred = as.factor(round(res3$net.result)), true = rw_test[,12])
ACC3[i]<-classAgreement(tab3,match.names=T)$diag
}

mean(ACC3);sd(ACC3)
mean(MSPE3); sd(MSPE3)

```

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}
dim(rw)
names(rw)
str(rw)
summary(rw)
#rw$alcohol_g<-rw$alcohol*rw$density*1000/100
```

Looking at the variables, it seems that medians and means are close so I expect symmetrical distributions for these observables.This will be checked using histograms below. The density spread is very narrow and I do not expect it to have a great influence on the quality of wine, but this also should be checked below. 

```{r echo=FALSE}
qplot(quality,data=rw)+ geom_histogram(colour="black", fill="darkred")+ xlab("QUALITY") + ylab("COUNT")+ theme(axis.text=element_text(size=16,color="black"))
```
```{r echo=FALSE}
ggplot(aes(quality,alcohol, color=quality),data=rw) + theme(legend.position="none")+ xlab("QUALITY") + ylab("ALCOHOL (% of vol)")+ theme(axis.text=element_text(size=16,color="black")) + geom_line(stat="summary", fun.y=median, size=2)+geom_point(position=position_jitter(h=0, w=0.4))
                                                                                                                                                                                 
ggplot(aes(quality,volatile.acidity, color=quality),data=rw)+geom_point(position=position_jitter())+ geom_line(stat="summary", fun.y=median, size=2)+ theme(legend.position="none")+ xlab("QUALITY") + ylab("VOLATILE ACIDITY (g/l)")+scale_y_continuous(limits=c(0.2, 1.2))+ theme(axis.text=element_text(size=16,color="black"))

ggplot(aes(quality,sulphates, color=quality),data=rw)+geom_point(position=position_jitter())+ geom_line(stat="summary", fun.y=median, size=2)+ theme(legend.position="none")+ xlab("QUALITY") + ylab("SULFATES (g/l)")+scale_y_continuous(limits=c(0.3, 1.2)) + theme(axis.text=element_text(size=16,color="black"))                                                                                                                                                                                                                                                                                                                                                                                     

ggplot(aes(quality,total.sulfur.dioxide, color=quality),data=rw)+geom_point(position=position_jitter())+ geom_line(stat="summary", fun.y=median,size=2)+ theme(legend.position="none")+ xlab("QUALITY") + ylab("TOTAL SULFUR DIOXIDE (mg/l)")+scale_y_continuous(limits=c(0, 160))+ theme(axis.text=element_text(size=16,color="black"))

ggplot(aes(quality,citric.acid, color=quality),data=rw)+geom_point(position=position_jitter())+ geom_line(stat="summary", fun.y=median, size=2) + theme(legend.position="none")+ xlab("QUALITY") + ylab("CITRIC ACID (g/l)")+scale_y_continuous(limits=c(0, 0.77))+ theme(axis.text=element_text(size=16,color="black"))

```

As expected this distribution is close to symmetric. Also, there are only discrete values of quality and most wines are 5 or 6 on the quality scale.

```{r echo=FALSE}
p1<-qplot(fixed.acidity,data=rw)
p2<-qplot(volatile.acidity,data=rw)
p3<-qplot(sqrt(citric.acid),data=rw)
grid.arrange(p1,p2,p3,nrow=1)
```

```{r echo=FALSE}
p4<-qplot(residual.sugar,data=rw)
p5<-qplot(chlorides,data=rw)
p6<-qplot(sqrt(free.sulfur.dioxide),data=rw)
grid.arrange(p4,p5,p6,nrow=1)
```

```{r echo=FALSE}
p7<-qplot(sqrt(total.sulfur.dioxide),data=rw)
p8<-qplot(density,data=rw)
p9<-qplot(pH,data=rw)
grid.arrange(p7,p8,p9,nrow=1)
```

```{r echo=FALSE}
p10<-qplot(sulphates,data=rw)
p11<-qplot(sqrt(alcohol),data=rw)
#p12<-qplot(compound.acidity,data=rw)
grid.arrange(p10,p11,nrow=1)
```
# Univariate Analysis

### What is the structure of your dataset?

The original data set had 1599 observables (different wines tasted) and 12 variables
(a number variable, 11 chemical properties plus a wine "quality" assessment). The variables are: "fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar", "chlorides",
"free.sulfur.dioxide", "total.sulfur.dioxide", "density", "pH", "sulphates", "alcohol" and "quality".
The median wine quality is 5.636 on a scale of 0 to 10. Most wines (1319) are given a quality rating of 5 or 6, the maximal
rating is 8.

### What is/are the main feature(s) of interest in your dataset?

The main feature is "quality" of wine and according to my research it depends mainly on the combination of sweetness and acidity. Hence, I expect it to depend mostly on the "alcohol", "residual.sugar" (these are the two variables that make wine taste "sweet"), and the variables 
for acidity: "fixed.acidity", "volatile.acidity", "citric.acid", and "pH". According to my research a high quality wine 
is not supposed to taste very sweet and acids counter-interact the sensation of sweetness from sugar and alcohol.

# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}
cor(rw[,12],rw[,1:11])
#ggpairs(data=rw[1:5,])
```

The variable quality is positively correlated most strongly with alcohol. The other variable for sweetness
residual.sugar has almost no effect on quality, probably because there are very small amounts of residual sugar, while alcohol is in much larger amounts per unit volume. As for the acidity variables, it appears that they do not have only positive or negative correlations as I theorized initially. Quality is also positively correlated with sulphates contrary to my expectations. It appears the sulphates in the amounts they are present do not contribute much to SO2 (which in large amounts spoils the 
taste in wine), but protect the quality of wine from microbial activity. The variable "pH" has a negligible correlation with quality, likely because it is more important which type of acid is predominant and not the total acidity of wine.

```{r echo=FALSE}
ggplot(aes(alcohol,quality),data=rw)+geom_point(position=position_jitter())
```

Generally, quality is growing with alcohol

```{r echo=FALSE}
ggplot(aes(alcohol,citric.acid),data=rw)+geom_boxplot()+facet_wrap(~quality)
```

```{r echo=FALSE}
ggplot(aes(quality, alcohol),data=rw)+geom_point(position=position_jitter()) + geom_line(stat="summary", fun.y=median)
```

```{r echo=FALSE}
ggplot(aes(quality, alcohol),data=rw) + geom_boxplot() + facet_wrap(~quality)
```

This is again a study of the relationship between alcohol and quality, faceted by quality. Generally there is a tendency
of the median alcohol content to grow (except for quality=5, where there is a dip). I will look more into this in the multivariate section.


# Bivariate Analysis

### What was the strongest relationship you found?

Quality strongly correlates with alcohol content.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(alcohol,volatile.acidity, alpha=1/2),data=rw)+geom_point(aes(color=quality))
ggplot(aes(volatile.acidity,quality, color=quality),data=rw)+geom_point(position=position_jitter()) 
```



# Reflection
 
The wine data set I worked with is not very big. Most wines (82%) are rated as of quality 5 or 6, yet many chemical properties are measured. The work with this set showed me how difficult it is to model human subjective experience such as how particular wine tastes. The alcohol quantity definitely influences the human taste the most, but at least partially it is balanced by the content of citric acid. It would have been interesting to have a variable for the brand of wine and see how this would influence the perceived wine quality. In addition I would have liked to see a year of tasting and correlate it with quality and weather for that particular year. With climate change generally weather is warmer and this may influence the quality of wine over time since in warmer climates the acidity of grapes is lower but the sugar content grows.  
